---
title: "Computing Assignment: Geo-data and Methods in R"
format: html
editor: visual
---

```{r}
#| label: packages
#| output: false
#| warning: false

# Load packages
library(here)
library(dismo)
library(rworldmap)
library(sf)
library(geodata)
library(tidyverse)
library(kableExtra)
library(gridExtra)
library(cols4all)

  
# Load world map
wrld_simpl <- getMap(resolution = "coarse")
```

#### Question 1 - Species 1 Distribution Modelling 

**Run linear models to predict the present-day distribution of species 1 using climate variables and use them to present a map of its current distribution. Which set of climatic variables best explain the current distribution of the species?**

```{r}
#| label: prepDataFunction
#| output: false
#| warning: false

prepData <- function(genus, species, rawFilename, processedFilename) {
  
  ### Downloading the data ###
  
  # Checks if the species data is already downloaded 
  if (!file.exists(here("data", "raw", rawFilename))) {
    # If not already downloaded, species data is downloaded from GBIF
    species.gbif <- gbif(genus, species, geo = TRUE)
    # The data is saved locally so it doesn't have to be downloaded again
    saveRDS(species.gbif, here("data", "raw", rawFilename))
  } else {
    # If the data is already downloaded the file is read
    species.gbif <- readRDS(here("data", "raw", rawFilename))
  }
  
  ### Cleaning the data ###
  
  species.coords <- species.gbif %>% 
    # Pull out the lat and lon columns
    transmute(lon, lat) %>% 
    # Remove empty rows 
    na.omit() %>% 
    # Ensure it's set to be a data frame
    as.data.frame()
  
  
  ### Remove the ocean areas from the map ###
  
  # Download ocean data
  ocean_data_dir <- here("data", "raw", "ocean")
  if (!dir.exists(ocean_data_dir)) dir.create(ocean_data_dir)
  URL <- "https://naturalearth.s3.amazonaws.com/110m_physical/ne_110m_ocean.zip"
  zip_file <- file.path(ocean_data_dir, basename(URL))
  if (!file.exists(zip_file)) {
    download.file(URL, zip_file)
  }
  
  # Unzip to ocean data directory and read shapefile
  files <- unzip(zip_file, exdir = ocean_data_dir)
  oceans <- read_sf(grep("shp$", files, value = TRUE))
  
  # Convert coordinates to a spatial features (sf) object for GIS operations
  species.coords <- st_as_sf(species.coords, coords = c("lon", "lat"))
  # Set the coordinate reference system (CRS) to match the oceans data
  st_crs(species.coords) <- st_crs(oceans)
  sf_use_s2(FALSE)  # Disable spherical geometry
  
  # Find where out points intersect with the ocean
  tmp <- sapply(st_intersects(species.coords, oceans), function(z) if (length(z) == 0) NA_integer_ else z[1])
  
  # Remove points that intersect with the ocean and convert back to table of coordinates
  if (sum(!is.na(tmp)) > 0) {
    species.coords <- data.frame(st_coordinates(species.coords[is.na(tmp), ]))
  } else {
    species.coords <- data.frame(st_coordinates(species.coords))
  }
  colnames(species.coords) <- c("lon", "lat")

  
  ### Extract climatic values for locations occupied by the species ###
  
  # Download bioclimatic data from the worldclim database and convert to Raster format
  bio.data <- worldclim_global(var = "bio", res = 10, path = here("data", "raw"))
  names(bio.data) <- paste0("bio", 1:19)
  
  # Extracting bioclimatic data for the focal localities where species is found
  bio.values <- terra::extract(bio.data, species.coords)[, -1]
  rownames(bio.values) <- rownames(species.coords)
  
  # Append to lat long, remove rows with missing data, and save to file for future use
  species.data <- cbind(species.coords, bio.values)
  write.csv(species.data, file = here("data", "processed", processedFilename), row.names = FALSE)
  
  
  ### Generate random background points for comparison in model ###
  
  # Define study extent based on species occurrence data (with some buffer around it)
  e <- extent(
    min(species.coords$lon) - 5,
    max(species.coords$lon) + 5,
    min(species.coords$lat) - 5,
    max(species.coords$lat) + 5
  )
  
  # Create a mask from the world map for the study region
  mask <- rasterize(wrld_simpl, raster(e, res = 0.5))
  
  # Generate 500 random background points within the study region
  bg <- randomPoints(mask, 500, ext = e)
  colnames(bg) <- c("lon", "lat")
  
  # Crop the bio.data to just keep values for this region
  bio.data <- crop(bio.data, e)
  
  
  ### Combine the presence data and the background data in one data frame ###
  
  train <- rbind(species.coords, bg)
  # Create a vector of 1s and 0s to indicate presence/absence
  pb_train <- c(rep(1, nrow(species.coords)), rep(0, nrow(bg)))
  # Extract the bioclimatic data for the presence and background points
  envtrain <- terra::extract(bio.data, train)
  envtrain <- data.frame(cbind(pa = pb_train, envtrain))
  # And for each set separately
  testpres <- data.frame(terra::extract(bio.data, species.coords))
  testbackg <- data.frame(terra::extract(bio.data, bg))
  
  # Saves the variables needed for the gm model as a list to be used later
  return(list(envtrain = envtrain, 
              testpres = testpres, 
              testbackg = testbackg, 
              bio.data = bio.data, 
              species.coords = species.coords,
              e = e))
}
```

```{r}
#| label: prepDataMegarhyssa
#| warning: false

megarhyssaPrep <- prepData(
  genus = "megarhyssa",
  species = "macrura ",
  rawFilename = "megarhyssa.gbif.rds",
  processedFilename = "megarhyssa.csv" 
)
```
```{r}
#| label: evaluateModelFuction

evaluateModel <- function(modelList, envtrain, testpres, testbackg) {
  
  # Set up data frame to store the results in 
  results <- data.frame(Model = character(), 
                        AUC = numeric(), 
                        AIC = numeric(), 
                        stringsAsFactors = FALSE)
  
  # Set up the for loop 
  for (name in names(modelList)) {
    # Create the model
    model <- glm(modelList[[name]], family = binomial(link = "logit"), data = envtrain)
    # Evaluate model
    eval <- evaluate(testpres, testbackg, model)
    # Extract AUC and AIC
    auc <- eval@auc
    # Extract AIC
    aic <- AIC(model)
    # Store results in data frame
    results <- rbind(results, data.frame(Model = name, AUC = auc, AIC = aic))
  }
  
  # Output the results
  return(results)
}

```

```{r initialModelsMergarhyssa}
initialList <- list("gm1" = pa ~ bio1, 
                    "gm2" = pa ~ bio2,
                    "gm3" = pa ~ bio3,
                    "gm4" = pa ~ bio4,
                    "gm5" = pa ~ bio5, 
                    "gm6" = pa ~ bio6,
                    "gm7" = pa ~ bio7,
                    "gm8" = pa ~ bio8,
                    "gm9" = pa ~ bio9,
                    "gm10" = pa ~ bio10)

initialModels <- evaluateModel(modelList = initialList,
                               envtrain = megarhyssaPrep$envtrain,
                               testpres = megarhyssaPrep$testpres,
                               testbackg = megarhyssaPrep$testbackg)

initialModels %>%
  kbl() %>%
  kable_styling(full_width = FALSE, position = "center")
```
```{r}
#| label: moreComplexModels

refinedList <- list("gm11"  = pa ~ bio1 + bio2, 
                    "gm12"  = pa ~ bio1 + bio5,
                    "gm13"  = pa ~ bio1 + bio10,
                    "gm14"  = pa ~ bio2 + bio5,
                    "gm15"  = pa ~ bio2 + bio10,
                    "gm16" = pa ~ bio5 + bio10,
                    "gm17" = pa ~ bio1 + bio2 + bio5,
                    "gm18" = pa ~ bio1 + bio2 + bio10,
                    "gm19" = pa ~ bio1 + bio5 + bio10,
                    "gm20" = pa ~ bio2 + bio5 + bio10,
                    "gm21" = pa ~ bio1 + bio2 + bio5 + bio10)

refinedModels <- evaluateModel(modelList = refinedList,
                               envtrain = megarhyssaPrep$envtrain,
                               testpres = megarhyssaPrep$testpres,
                               testbackg = megarhyssaPrep$testbackg)

refinedModels %>%
  kbl() %>%
  kable_styling(full_width = FALSE, position = "center")
```

```{r}
#| label: plotModelFunction 

plottingModel <- function(formula, bio.data, species.coords, envtrain, testpres, testbackg, e){
  # Model 
  gm <- glm(formula,
            family = binomial(link = "logit"), 
            data = envtrain)
  
  # Predict pg, a raster layer representing the probability of occurance from the linear model for area of     extent e
  pg <- predict(bio.data, gm, ext = e, type = "response")
  pg <- crop(pg, e)
  
  # Set up so the plots will go side by side
  par(mfrow = c(2,1))
  
  # Plot this probability layer
  print(plot(pg, main = "GLM probability of occurrence"))
  # Add country boundaries
  print(plot(wrld_simpl, add = TRUE, border = "dark grey"))
  # Add our observed locality data
  print(points(species.coords, col="red", pch=20, cex=1.5))
  
  # Evaluates how well the model predicts presence/absence at each point
  ge <- evaluate(testpres, testbackg, gm)
  print(ge)
  
  # This evaluation picks a threshold probability for defining presence/absence
  tr <- threshold(ge, "prevalence")
  print(plot(pg > tr, main = "presence/absence"))
  print(plot(wrld_simpl, add = TRUE, border = "dark grey"))
  print(points(species.coords, col="red", pch = 20, cex = 1.5))
}
```

```{r}
#| label: plotMegarhyssa
#| results: "hide"
#| fig-width: 8
#| fig-height: 12

plottingModel(formula = pa ~ bio1 + bio2 + bio5 + bio10,
              bio.data = megarhyssaPrep$bio.data,
              species.coords = megarhyssaPrep$species.coords,
              envtrain = megarhyssaPrep$envtrain,
              testpres = megarhyssaPrep$testpres,
              testbackg = megarhyssaPrep$testbackg,
              e = megarhyssaPrep$e)
```

#### Question 2 - Species 2 Distribution Modelling 

**Run linear models to predict the present-day distribution of species 2 using climate variables and use them to present a map of its current distribution. Which set of climatic variables best explain the current distribution of the species?**

```{r}
#| label: prepDataTremex

tremexPrep <- prepData(
  genus = "tremex",
  species = "columba",
  rawFilename = "tremex.gbif.rds",
  processedFilename = "tremex.csv" 
)
```
```{r initialModelsTremex}
initialList <- list("gm1" = pa ~ bio1, 
                    "gm2" = pa ~ bio2,
                    "gm3" = pa ~ bio3,
                    "gm4" = pa ~ bio4,
                    "gm5" = pa ~ bio5, 
                    "gm6" = pa ~ bio6,
                    "gm7" = pa ~ bio7,
                    "gm8" = pa ~ bio8,
                    "gm9" = pa ~ bio9,
                    "gm10" = pa ~ bio10)

initialModels <- evaluateModel(modelList = initialList,
                               envtrain = tremexPrep$envtrain,
                               testpres = tremexPrep$testpres,
                               testbackg = tremexPrep$testbackg)

initialModels %>%
  kbl() %>%
  kable_styling(full_width = FALSE, position = "center")
```
```{r}
#| label: refinedTremexModels

refinedList <- list("gm11" = pa ~ bio2 + bio3, 
                    "gm12" = pa ~ bio2 + bio7,
                    "gm13" = pa ~ bio2 + bio10,
                    "gm14" = pa ~ bio3 + bio7,
                    "gm15" = pa ~ bio3 + bio10, 
                    "gm16" = pa ~ bio7 + bio10,
                    "gm17" = pa ~ bio2 + bio3 + bio7,
                    "gm18" = pa ~ bio2 + bio3 + bio10,
                    "gm19" = pa ~ bio2 + bio7 + bio10,
                    "gm20" = pa ~ bio3 + bio7 + bio10,
                    "gm21" = pa ~ bio2 + bio3 + bio7 + bio10)

refinedModels <- evaluateModel(modelList = refinedList,
                               envtrain = tremexPrep$envtrain,
                               testpres = tremexPrep$testpres,
                               testbackg = tremexPrep$testbackg)

refinedModels %>%
  kbl() %>%
  kable_styling(full_width = FALSE, position = "center")
```
```{r}
#| label: plotTremex
#| results: "hide"
#| fig-width: 8
#| fig-height: 12

plottingModel(formula = pa ~ bio2 + bio3 + bio7 + bio10,
              bio.data = tremexPrep$bio.data,
              species.coords = tremexPrep$species.coords,
              envtrain = tremexPrep$envtrain,
              testpres = tremexPrep$testpres,
              testbackg = tremexPrep$testbackg,
              e = tremexPrep$e)
```


#### Question 3 - Distribution Overlap

**Plot the overlap in distribution of the two species. Devise and calculate a metric for the degree of overlap between their ranges, explaining how you calculated it.**

```{r}
### First species ###

# Runs the linear model
gm1 <- glm(pa ~ bio1 + bio2 + bio5 + bio10,
          family = binomial(link = "logit"), 
          data = megarhyssaPrep$envtrain)

# Predict pg, a raster layer representing the probability of occurance 
# from the linear model for area of extent e
pg1 <- predict(megarhyssaPrep$bio.data, 
               gm1, 
               ext = megarhyssaPrep$e, 
               type = "response")

# Crop pg to e
pg1 <- crop(pg1, megarhyssaPrep$e)

# Evaluates how well the model predicts presence/absence at each point
ge1 <- evaluate(megarhyssaPrep$testpres, megarhyssaPrep$testbackg, gm1)

# This evaluation picks a threshold probability for defining presence/absence
tr1 <- threshold(ge1, "prevalence")

# Saves where the model predicts the species to be present
presence1 <- pg1 > tr1
presence1 <- mask(presence1, presence1, maskvalue = FALSE)

### Second species ###

# Runs the linear model 
gm2 <- glm(pa ~ bio2 + bio3 + bio7 + bio10,
           family = binomial(link = "logit"), 
           data = tremexPrep$envtrain)

# Predict pg, a raster layer representing the probability of occurance 
# from the linear model for area of extent e
pg2 <- predict(tremexPrep$bio.data, 
               gm2, 
               ext = megarhyssaPrep$e, 
               type = "response")

# Crop pg to e
pg2 <- crop(pg2, megarhyssaPrep$e)

# Evaluates how well the model predicts presence/absence at each point
ge2 <- evaluate(tremexPrep$testpres, tremexPrep$testbackg, gm2)

# This evaluation picks a threshold probability for defining presence/absence
tr2 <- threshold(ge2, "prevalence")

# Saves where the model predicts the species to be present
presence2 <- pg2 > tr2
presence2 <- mask(presence2, presence2, maskvalue = FALSE)

### Overlap ###

# Identify areas where both species are predicted to be present
overlap <- presence1 & presence2

# Mask to remove FALSE values
overlap <- mask(overlap, overlap, maskvalue = FALSE)

plot(overlap)
plot(wrld_simpl, add = TRUE, border = "dark grey")


### Area calculations ###

area(overlap)

area1 <- expanse(presence1, unit = "km")
area2 <- expanse(presence2, unit = "km")
areaOverlap <- expanse(overlap, unit = "km")

overlapMetric = areaOverlap / ((area1 + area2) - areaOverlap)

overlapMetric

### Plotting ###

# Sorting out where the map is 
> wrld_simpl <- vect(wrld_simpl)  # Convert to terra's vector format
> wrld_simpl <- crop(wrld_simpl, presence1) 

# Plotting the two predicted species distributions, and the predicted overlap
plot(presence1, col = "#88CCEE")
plot(presence2, col = "#CC6677", add = TRUE)
plot(overlap, col = "#332288", add = TRUE)
plot(wrld_simpl, add = TRUE, border = "dark grey")
```

#### Question 4 - Interspecies Distribution Dependence

**Use a linear model to test whether the distribution of species 1 at the present time depends on the distribution of species 2, while also taking account of the effects of climatic variables.**